(Deep) Induction for GADTs

### 2021-03-30 induction-paper/GADT-induction.tex
Rearranged material in section 4

@@

In this section we will generalize the approach taken in the previous examples
and provide a general framework to derive induction rules for GADTs.
We restrict our scope to GADTs not featuring nesting,
in the sense that their constructors contain no nested calls of the recursive variable,
as truly nested types do.
The reason for this restriction will be discussed in Section~\ref{sec:GADT-nested}.

Consider a generic GADT
\begin{equation}\label{eq:gadt}
\begin{array}{l}
\mathsf{data\, G\,(\ol{a : Set}) : Set\,where}\\
\mathsf{\;\;\;\;\;\;\;\;C\, :\, F\,G\,\ol{b} \to G (\ol{K\,\ol{b}})}
\end{array}
\end{equation}
which, for ease of notation,
we assume to have only one constructor $\mathsf{C}$.
Extending the argument to the case of a GADT with multiple constructors presents no difficulty.
In the definition above, $\mathsf{F}$ has type $\mathsf{(Set^{\alpha} \to Set) \to Set^{\beta} \to Set}$
and each $\mathsf{K}$ has type $\mathsf{Set^{\beta} \to Set}$.
The overline notation denotes a finite list:
%and the vertical bars notation denotes the length of a list:
$\mathsf{G}$ depends on a finite list of types $\mathsf{\ol{a}}$ of length $\mathsf{\alpha}$,
each $\mathsf{K}$ depends on a finite list of types $\mathsf{\ol{b}}$ of length $\mathsf{\beta}$,
and notice that the length of the list $\mathsf{\ol{a}}$ matches that of the list $\mathsf{\ol{K}}$.
With that in mind, the constructor $\mathsf{C}$ can be rewritten as
\[
\mathsf{C : \exists (\ol{b : Set}) \to \ol{Equal\,a\,(K\,\ol{b})} \to F\,G\,\ol{b} \to G\,\ol{a}}
\]
which is the form we shall use from now on.

In order to state the induction rule for $\mathsf{G}$,
we first need to define its associated predicate-lifting function 
\[
\mathsf{G^{\wedge} : \forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set}
\]
as
\[
\mathsf{G^{\wedge}\,\ol{a}\,\ol{Q_a}\,(C\,\ol{b}\,\ol{e_a}\,x)
= \exists (\ol{Q_b : b \to Set})
\to \ol{Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}
\times F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,{\ol{Q_b}}\,x}
\]
where $\mathsf{Q_a : a \to Set}$, $\mathsf{e_a : Equal\,a\,(K\,\ol{b})}$ and $\mathsf{x : F\,G\,\ol{b}}$.
Notice that we need a type $\mathsf{Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}$
for each type $\mathsf{a}$.
We also assume to have liftings for $\mathsf{F}$,
\begin{multline*}
\mathsf{F^{\wedge} : \forall (G : Set^{\alpha} \to Set) (\ol{b : Set})
\to (\forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set)} \\
\mathsf{\to (\ol{b \to Set})
\to F\,G\,\ol{b} \to Set}
\end{multline*}
and for each  $\mathsf{K}$,
\[
\mathsf{K^{\wedge} : \forall (\ol{b : Set}) \to (\ol{b \to Set}) \to K\,\ol{b} \to Set}
\]
as we could not possibly define an induction rule otherwise.
A way to concretely define the predicate-lifting function for a type is to proceed by induction on the structure of the type.
% on the structure of $\mathsf{F}$ and $\mathsf{K}$ respectively.

@@

Define the induction rule for $\mathsf{G}$

@@

\mathsf{\to \forall (\ol{a : Set})(\ol{Q_a : a \to Set})(x : G\,\ol{a}) \to G^{\wedge}\,\ol{a}\,\ol{Q_a}\,x \to P\,\ol{a}\,\ol{Q_a}\,x}

@@

%$\mathsf{\ol{pc_i : PC_i}}$,
%$\mathsf{\ol{Q_a : a \to Set}}$,
we have $\mathsf{e_a : Equal\,a\,(K\,\ol{b})}$ for each $\mathsf{a}$,
$\mathsf{x : F\,G\,\ol{b}}$,
$\mathsf{Q_b : b \to Set}$ for each $\mathsf{b}$ ,

@@

$\mathsf{L_F : F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,{\ol{Q_b}}\,x}$,
and

@@

\mathsf{F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x = \forall (y : F_1\,G\,\ol{b}) \to F_2^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,(x\,y)} \\

@@

\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x = \forall (y : F_1\,G\,\ol{b}) \to F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,(x\,y)}

@@

@@

Thus, we define $\mathsf{p\,x\,L_F\,y = p_2 (x\,y) (L_F\,y)}$.

@@

Notice that the fact that the types $\mathsf{F_a}$ do not contain the recursive variable

@@

As mentioned before, that is a fundamental assumption to make the argument work.
Why the argument fails in presence of truly nested GADTs will be shown in the next section.

@@

\item Case $\mathsf{F\,G\,\ol{b} = H\,(\ol{F_c\,G\,\ol{b}})}$

@@

That means that $\mathsf{H}$ cannot be a GADT, as those do not have functorial semantics~\cite{HaskellPaper},

@@

%\subsection{General case}
%%Outline general heuristics for this process, generalizing process in deep induction paper.
%%\begin{itemize}
%%    \item Lifting for GADT (possibly a restricted class?)
%%    \item Deep induction rule
%%    \item Prove deep induction (define term)
%%\end{itemize}
%TODO:
%\begin{itemize}
%\item explain overline is vector
%\end{itemize}
%In this section we will generalize the approach taken in the previous examples
%and provide a general framework to derive induction rules for GADTs.
%We restrict our scope to GADTs not featuring nesting,
%in the sense that their constructors contain no nested calls of the recursive variable,
%as truly nested types do.
%The reason for this restriction will be discussed in Section~\ref{sec:GADT-nested}.
%Consider a generic GADT
%\begin{equation}\label{eq:gadt}
%\begin{array}{l}
%\mathsf{data\, G\,(\ol{a : Set}) : Set\,where}\\
%\mathsf{\;\;\;\;\;\;\;\;C_i\, :\, F_i\,G\,\ol{b} \to G (\ol{K_i\,\ol{b}})}
%\end{array}
%\end{equation}
%with a finite number of constructors $\mathsf{C_i}$,
%and where $\mathsf{F_i}$ has type $\mathsf{(Set^{\lvert\ol{a}\rvert} \to Set) \to Set^{\lvert\ol{b}\rvert} \to Set}$
%and each $\mathsf{K_i}$ has type $\mathsf{Set^{\lvert\ol{b}\rvert} \to Set}$.
%The overline notation denotes a finite list:
%$\mathsf{G}$ depends on a finite list of types $\mathsf{a}$,
%each $\mathsf{K_i}$ depends on a finite list of types $\mathsf{b}$,
%and notice that the length of the list of $\mathsf{a}$ matches that of the list of $\mathsf{K_i}$.
%With that in mind, each constructor $\mathsf{C_i}$ can be rewritten as
%\mathsf{C_i : \exists (\ol{b : Set}) \to \ol{Equal\,a\,(K_i\,\ol{b})} \to F_i\,G\,\ol{b} \to G\,\ol{a}}
%which is the form we shall use from now on.
%We define the lifting of $\mathsf{G}$
%\mathsf{G^{\wedge} : \forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set}
%\mathsf{G^{\wedge}\,\ol{a}\,\ol{Q_a}\,(C_i\,\ol{b}\,\ol{e_a}\,x)
%= \exists (\ol{Q_b : b \to Set})
%\to \ol{Equal^{\wedge}\,a\,(K_i\,\ol{b})\,Q_a\,(K_i^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}
%\times F_i^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,{\ol{Q_b}}\,x}
%where $\mathsf{Q_a : a \to Set}$, $\mathsf{e_a : Equal\,a\,(K_i\,\ol{b})}$ and $\mathsf{x : F_i\,G\,\ol{b}}$,
%and the liftings
%\mathsf{F_i^{\wedge} : \forall (G : Set^{\lvert\ol{a}\rvert} \to Set) (\ol{b : Set})
%\to (\forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set)} \\
%\mathsf{\to (\ol{b \to Set})
%\to F_i\,G\,\ol{b} \to Set}
%\mathsf{K_i^{\wedge} : \forall (\ol{b : Set}) \to (\ol{b \to Set}) \to K_i\,\ol{b} \to Set}
%are defined by induction on the structure of the type.
%Associate the type
%\mathsf{PC_i = \forall (\ol{a : Set}) (\ol{b : Set}) (\ol{Q_a : a \to Set}) (\ol{Q_b : b \to Set}) (\ol{e_a : Equal\,a\,(K_i\,\ol{b})}) (x : F_i\,G\,\ol{b})} \\
%\mathsf{\to \ol{Equal^{\wedge}\,a\,(K_i\,\ol{b})\,Q_a\,(K_i^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}
%	\to F_i^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x
%	\to P\,\ol{a}\,\ol{Q_a}\,(C_i\,\ol{b}\,\ol{e_a}\,x)}
%to each constructor $\mathsf{C_i}$.
%Define the induction rule for $\mathsf{G}$
%\mathsf{DIG : \forall (P : \forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set)} \\
%\mathsf{\to \bigtimes_i PC_i
%\to \forall (\ol{a : Set})(\ol{Q_a : a \to Set})(x : G\,\ol{a}) \to G^{\wedge}\,\ol{a}\,\ol{Q_a}\,x \to P\,\ol{a}\,\ol{Q_a}\,x}
%\mathsf{DIG\,P\,\ol{pc_i}\,\ol{a}\,\ol{Q_a}\,(C_i\,\ol{b}\,\ol{e_a}\,x)\,(\ol{Q_b}, \ol{E_a}, L_F)
%= pc_i\,\ol{a}\,\ol{b}\,\ol{Q_a}\,\ol{Q_b}\,\ol{e_a}\,x\,\ol{E_a}\,(p\,L_F)}
%where $\mathsf{\ol{pc_i : PC_i}}$,
%$\mathsf{\ol{Q_a : a \to Set}}$,
%$\mathsf{\ol{e_a : Equal\,a\,(K_i\,\ol{b})}}$,
%$\mathsf{x : F_i\,G\,\ol{b}}$,
%$\mathsf{\ol{Q_b : b \to Set}}$,
%$\mathsf{\ol{E_a : Equal^{\wedge}\,a\,(K_i\,\ol{b})\,Q_a\,(K_i^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}}$,
%$\mathsf{L_F : F_i^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,{\ol{Q_b}}\,x}$,
%and $\mathsf{p}$ is defined in the following lemma.
%\begin{lemma}
%For every $\mathsf{F, G, \ol{b}, P, \ol{Q_b}, x}$, there exists a function
%\mathsf{p : F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x \to F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x}
%\end{lemma}
%\begin{proof}
%By structural induction on $\mathsf{F}$:
%\begin{itemize}
%\item Case $\mathsf{F = F_1 \times F_2}$.
%We have that $\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x = F_1^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x \times F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x}$.
%By inductive hypothesis, there exist functions
%$\mathsf{p_1 : F_1^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x_1 \to F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x_1}$ and
%$\mathsf{p_2 : F_2^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x_2 \to F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x_2}$.
%Thus, we define $\mathsf{p (L_1, L_2) = (p_1\,L_1) \times (p_2\,L_2)}$
%%%%
%\item Case $\mathsf{F = F_1 + F_2}$. Analogous to case $\mathsf{F = F_1 \times F_2}$.
%%%%
%\item Case $\mathsf{F = F_1 \to F_2}$ where $\mathsf{F_1}$ does not contain the recursive variable.
%We have that
%\mathsf{F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x = F_1^{\wedge}\,\ol{b}\,\ol{Q_b}\,x \to F_2^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x} \\
%By inductive hypothesis, there exist a function
%$\mathsf{p_2 : F_2^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x_2 \to F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x_2}$.
%Thus, we define $\mathsf{p\,L\,l = p_2\,(L\,l)}$ for
%$\mathsf{L : F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x}$
%and $\mathsf{l : F_1^{\wedge}\,\ol{b}\,\ol{Q_b}\,x}$.
%{\color{blue} TODO: explain why $\mathsf{F_1}$ does not contain the recursive variable.}
%%%%
%\item Case $\mathsf{F\,G\,\ol{b} = G(\ol{F_a\,\ol{b}})}$.
%Thus, $\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b} = P (\ol{F_a\,\ol{b}}) (\ol{F_a^{\wedge}\,\ol{b}\,\ol{Q_b}})}$.
%So, $\mathsf{p}$ is defined as
%\mathsf{p\,L_F = DIG\,P\,\ol{pc_i}\,(\ol{F_a\,\ol{b}})\,(\ol{F_a^{\wedge}\,\ol{b}\,\ol{Q_b}})\,x\,L_F}
%Note: No nesting!
%\end{itemize}
%\end{proof}

%Define $\mathsf{p_b}$ by structural induction on $F_i$.
%
%Case $\mathsf{F_i\,G\,\ol{b} = F'_i\,G\,\ol{b} \times F''_i\,G\,\ol{b}}$.
%We have $\mathsf{F_i^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,(x', x'') = (F'_i)^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x' \times (F''_i)^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x''}$.
%Thus define
%\[
%\mathsf{
%p_b\,P\,PC_i\,\ol{a}\,\ol{Q_a}\,\ol{b}\,\ol{e_a}\,(x', x'')\,\ol{e'_a}\,(l', l'')
%=
%(
%p'_b\,P\,PC_i\,\ol{a}\,\ol{Q_a}\,\ol{b}\,\ol{e_a}\,x'\,\ol{e'_a}\,l',
%p''_b\,P\,PC_i\,\ol{a}\,\ol{Q_a}\,\ol{b}\,\ol{e_a}\,x''\,\ol{e'_a}\,l''
%)
%} 
%\]
%where $\mathsf{p'_b}$ is the function associated to $\mathsf{F'_i}$
%and $\mathsf{p''_b}$ is the function associated to $\mathsf{F''_i}$.
%
%
%Case $\mathsf{F_i\,G\,\ol{b} = H(\ol{F'_i\,G\,\ol{b}})}$ with $\mathsf{H}$ data type with constructors $\mathsf{D_i}$.
%We have
%\[
%\mathsf{F_i^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x = H^{\wedge}\,(\ol{F'_i\,G\,\ol{b}}) (\ol{(F'_i)^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}}) x}
%\]
%Thus define
%\[
%\mathsf{
%p_b\,P\,PC_i\,\ol{a}\,\ol{Q_a}\,\ol{b}\,\ol{e_a}\,x\,\ol{e'_a}\,l
%=
%} 
%\]

>>>>

\subsection{Induction for GADTs with nesting}\label{sec:GADT-nested}

In the previous sections,
we derive induction rules for examples of GADTs that do not feature nesting,
in the sense that their constructors contain no nested calls of the recursive variable,
as truly nested types do.
Since both nested types~\cite{jp20}
and GADTs without nesting admit induction rules,
it is just natural to expect that GADTs with nesting would as well.
Surprisingly, that is not the case:
indeed, induction rules for nested types rely on functorial semantics,
but GADTs cannot admit both functorial and parametric semantics at the same time~\cite{HaskellPaper}.
In this section we show how induction for GADTs featuring nesting goes wrong
by analyzing the following concrete example.
\begin{equation}
\begin{array}{l}
\mathsf{data\, G\,(a : Set) : Set\,where}\\
\mathsf{\;\;\;\;\;\;C :\, G(G\,a) \to G(a \times a)}
\end{array}
\end{equation}
The constructor $\mathsf{C}$ can be rewritten as
\[
\mathsf{C : \exists (b : Set) \to Equal\,a\,(b \times b) \to G(G\,b) \to G\,a}
\]
which is the form we shall use from now on.
Its predicate-lifting function
\[
\mathsf{G^{\wedge} : \forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set}
\]
is defined as
\[
\mathsf{G^{\wedge}\,a\,Q_a\,(C\,b\,e\,x)
= \exists (Q_b : b \to Set)
\to Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e
\times G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,x}
\]
where $\mathsf{Q_a : a \to Set}$, $\mathsf{e : Equal\,a\,(b \times b)}$ and $\mathsf{x : G(G\,b)}$.

An induction rule for $\mathsf{G}$ is a term
\begin{multline}
\mathsf{DIG : \forall (P : \forall (a : Set) \to (a \to Set) \to G\,a \to Set)} \\
\mathsf{\to \Big( \forall (a\;b : Set) (Q_a : a \to Set) (Q_b : b \to Set) (e : Equal\,a\,(b \times b)) (x : G\,(G\,b))} \\
\mathsf{\to Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e
	\to P\,(G\,b)\,(P\,b\,Q_b)\,x
	\to P\,a\,Q_a\,(C\,b\,e\,x) \Big)}\\
\mathsf{\to \forall (a : Set)(Q_a : a \to Set)(x : G\,a) \to G^{\wedge}\,a\,Q_a\,x \to P\,a\,Q_a\,x}
\end{multline}
Consistently with the examples in the previous sections, we define $\mathsf{DIG}$ as
\[
\mathsf{DIG\,P\,pc\,a\,Q_a\,(C\,b\,e\,x)\,(Q_b, L_E, L_G)
= pc\,a\,b\,Q_a\,Q_b\,e\,x\,L_E\,p}
\]
where $\mathsf{C\,b\,e\,x : G\,a}$,
%, the second to last term applied to $\mathsf{DIG}$,
%has type $\mathsf{G\,a}$,
i.e., $\mathsf{e : Equal\,a\,(b \times b)}$ and $\mathsf{x : G(G\,b)}$,
and $\mathsf{(Q_b, L_E, L_G) : G^{\wedge}\,a\,Q_a\,(C\,b\,e\,x)}$,
%, the last term applied to $\mathsf{DIG}$,
%has type $\mathsf{G^{\wedge}\,a\,Q_a\,(C\,b\,e\,x)}$,
i.e., $\mathsf{Q_b : b \to Set}$, $\mathsf{L_E : Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e}$,
and $\mathsf{L_G : G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,x}$.
%\begin{gather*}
%\mathsf{Q_b : b \to Set} \\
%\mathsf{L_E : Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e} \\
%\mathsf{L_G : G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,x}
%%\mathsf{p = DIG\,P\,pc\,(G\,b)\,(P\,b\,Q_b)\,x\,q : P\,(G\,b)\,(P\,b\,Q_b)\,x}
%\end{gather*}
We still need to define $\mathsf{p : P\,(G\,b)\,(P\,b\,Q_b)\,x}$.
We do so by using the induction rule and letting
\[
\mathsf{p = DIG\,P\,pc\,(G\,b)\,(P\,b\,Q_b)\,x\,q}
\]
where we still need to provide $\mathsf{q : G^{\wedge}\,(G\,b)\,(P\,b\,Q_b)\,x}$.
To produce such $\mathsf{q}$, we need the map function of $\mathsf{G^{\wedge}}$, with type
\[
\mathsf{GLMap : \forall (a : Set) (Q_a\;Q_a' : a \to Set) \to PredMap\,a\,Q_a\,Q_a' \to PredMap\,(G\,a)\,(G^{\wedge}\,a\,Q_a)\,(G^{\wedge}\,a\,Q_a')}
\]
where $\mathsf{PredMap : \forall (a : Set) \to (a \to Set) \to (a \to Set) \to Set}$
is defined as
\[
\mathsf{PredMap\,a\,Q_a\,Q_a' = \forall (x : a) \to Q_a\,x \to Q_a'\,x}
\]
and represents the type of morphisms between predicates.
If we had $\mathsf{GLMap}$, then we would be able to define
\[
\mathsf{q = GLMap\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,(P\,b\,Q_b)\,(DIG\,P\,pc\,b\,Q_b)\,x\,L_G}
\]
Unfortunately, we cannot define such a $\mathsf{GLMap}$.
Indeed, its definition would have to be
\[
\mathsf{GLMap\,a\,Q_a\,Q_a'\,M\,(C\,b\,e\,x)\,(Q_b, L_E, L_G) = (Q_b', L_E', L_G')}
\]
where $\mathsf{(Q_b, L_E, L_G)}$ has type $\mathsf{G^{\wedge}\,a\,Q_a\,(C\,b\,e\,x)}$
and $\mathsf{(Q_b', L_E', L_G')}$ has type $\mathsf{G^{\wedge}\,a\,Q_a'\,(C\,b\,e\,x)}$, i.e.,
\begin{gather*}
\mathsf{Q_a\;Q_a' : a \to Set} \\
\mathsf{Q_b\;Q_b' : b \to Set} \\
%\mathsf{M : PredMap\,b\,Q_a\,Q_a'} \\
\mathsf{L_E : Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e} \\
\mathsf{L_G\ : G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,x} \\
\mathsf{L_E' : Equal^{\wedge}\,a\,(b \times b)\,Q_a'\,(Pair^{\wedge}\,b\,b\,Q'_b\,Q'_b)\,e} \\
\mathsf{L_G' : G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b')\,x}
\end{gather*}
In other words, we have a proof $\mathsf{L_E}$ of the (extensional) equality
of the predicates $\mathsf{Q_a}$ and $\mathsf{Pair^{\wedge}\,b\,b\,Q_b\,Q_b}$
and a morphism of predicates $\mathsf{M}$ from $\mathsf{Q_a}$ to $\mathsf{Q_a'}$,
and we need to use those to deduce a proof of the (extensional) equality
of the predicates $\mathsf{Q_a'}$ and $\mathsf{Pair^{\wedge}\,b\,b\,Q_b'\,Q_b'}$.
But that is not generally possible:
the facts that $\mathsf{Q_a}$ is equal to $\mathsf{Pair^{\wedge}\,b\,b\,Q_b\,Q_b}$
and that there is a morphism of predicates $\mathsf{M}$ from $\mathsf{Q_a}$ to $\mathsf{Q_a'}$
do not guarantee that $\mathsf{Q_a'}$ is equal to $\mathsf{Pair^{\wedge}\,bb\,Q_b'\,Q_b'}$ for some $\mathsf{Q_b'}$.

@@

%We restrict our scope to GADTs not featuring nesting,
%in the sense that their constructors contain no nested calls of the recursive variable,
%as truly nested types do.
%This way we can avoid the issue described in Section~\ref{sec:GADT-nested}.
%The reason for this restriction will be discussed in Section~\ref{sec:GADT-nested}.

In this section we generalize the approach taken in the previous examples
and provide a general framework to derive induction rules for GADTs.
Since, as seen in Section~\ref{sec:GADT-nested},
GADTs featuring nesting do not admit an induction rule,
we restrict our scope to GADTs not featuring nesting.
Consider such a generic GADT
\begin{equation}\label{eq:gadt}
\begin{array}{l}
\mathsf{data\, G\,(\ol{a : Set}) : Set\,where}\\
\mathsf{\;\;\;\;\;\;\;\;C\, :\, F\,G\,\ol{b} \to G (\ol{K\,\ol{b}})}
\end{array}
\end{equation}
which, for ease of notation,
we assume to have only one constructor $\mathsf{C}$.
Extending the argument to GADTs with multiple constructors presents no difficulty.
In the definition above, $\mathsf{F}$ has type $\mathsf{(Set^{\alpha} \to Set) \to Set^{\beta} \to Set}$
and each $\mathsf{K}$ has type $\mathsf{Set^{\beta} \to Set}$.
The overline notation denotes a finite list:
%and the vertical bars notation denotes the length of a list:
$\mathsf{G}$ depends on a finite list of types $\mathsf{\ol{a}}$ of length $\mathsf{\alpha}$,
each of the $\mathsf{\alpha}$-many $\mathsf{K}$ depends on a finite list of types $\mathsf{\ol{b}}$ of length $\mathsf{\beta}$,
and notice that the length of the list of types $\mathsf{\ol{a}}$ matches that of the list of type constructors $\mathsf{\ol{K}}$.
With that in mind, the constructor $\mathsf{C}$ can be rewritten as
\[
\mathsf{C : \exists (\ol{b : Set}) \to \ol{Equal\,a\,(K\,\ol{b})} \to F\,G\,\ol{b} \to G\,\ol{a}}
\]
which is the form we shall use from now on.
Notice that, in the above constructor, we have a type
$\mathsf{Equal\,a\,(K\,\ol{b})}$ for each $\mathsf{a}$ and its corresponding type constructor $\mathsf{K}$.

In order to state the induction rule for $\mathsf{G}$,
we first need to define $\mathsf{G}$'s associated predicate-lifting function 
\[
\mathsf{G^{\wedge} : \forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set}
\]
as
\[
\mathsf{G^{\wedge}\,\ol{a}\,\ol{Q_a}\,(C\,\ol{b}\,\ol{e_a}\,x)
= \exists (\ol{Q_b : b \to Set})
\to \ol{Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}
\times F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,{\ol{Q_b}}\,x}
\]
where $\mathsf{Q_a : a \to Set}$
and $\mathsf{C\,\ol{b}\,\ol{e_a}\,x : G\,\ol{a}}$,
i.e., $\mathsf{e_a : Equal\,a\,(K\,\ol{b})}$ and $\mathsf{x : F\,G\,\ol{b}}$.
Notice that we need a type $\mathsf{Equal^{\wedge}\,a\,(K\,\ol{b})\,Q_a\,(K^{\wedge}\,\ol{b}\,\ol{Q_b})\,e_a}$
for each type $\mathsf{a}$ and its corresponding type constructor $\mathsf{K}$.
We also assume to have liftings for $\mathsf{F}$,
\begin{multline*}
\mathsf{F^{\wedge} : \forall (G : Set^{\alpha} \to Set) (\ol{b : Set})
\to (\forall (\ol{a : Set}) \to (\ol{a \to Set}) \to G\,\ol{a} \to Set)} \\
\mathsf{\to (\ol{b \to Set})
\to F\,G\,\ol{b} \to Set}
\end{multline*}
and for each  $\mathsf{K}$,
\[
\mathsf{K^{\wedge} : \forall (\ol{b : Set}) \to (\ol{b \to Set}) \to K\,\ol{b} \to Set}
\]
as we could not possibly define an induction rule otherwise.
A way to concretely define the predicate-lifting function for a type is to proceed by induction on the structure of the type,
and we have seen in the previous sections examples of how to do so for products and type application.
We do not give here the general definition of lifting,
as that would require to first present a full type calculus,
and that is beyond the scope of the paper.

@@

We can now define the induction rule for $\mathsf{G}$

@@

\mathsf{\to \forall (\ol{a : Set})(\ol{Q_a : a \to Set})(y : G\,\ol{a}) \to G^{\wedge}\,\ol{a}\,\ol{Q_a}\,y \to P\,\ol{a}\,\ol{Q_a}\,y}

@@

\item $\mathsf{C\,\ol{b}\,\ol{e_a}\,x : G\,\ol{a}}$,
i.e., $\mathsf{e_a : Equal\,a\,(K\,\ol{b})}$ for each $\mathsf{a}$
and $\mathsf{x : F\,G\,\ol{b}}$;
\item $\mathsf{(\ol{Q_b}, \ol{E_a}, L_F) : G^{\wedge}\,\ol{a}\,\ol{Q_a} (C\,\ol{b}\,\ol{e_a}\,x)}$,
i.e.,
$\mathsf{Q_b : b \to Set}$ for each $\mathsf{b}$,

@@

and $\mathsf{L_F : F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,{\ol{Q_b}}\,x}$.
\end{itemize}
Finally, the morphism of predicates

@@

\mathsf{F^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,x = \forall (z : F_1\,G\,\ol{b}) \to F_2^{\wedge}\,G\,\ol{b}\,G^{\wedge}\,\ol{Q_b}\,(x\,y)} \\

@@

\mathsf{F^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,x = \forall (z : F_1\,G\,\ol{b}) \to F_2^{\wedge}\,G\,\ol{b}\,P\,\ol{Q_b}\,(x\,y)}

@@

where $\mathsf{x : F\,G\,\ol{b} = F_1\,G\,\ol{b} \to F_2\,G\,\ol{b}}$.

@@

Thus, we define $\mathsf{p\,x\,L_F\,z = p_2 (x\,z) (L_F\,z)}$.

@@

Note that the fact that the type constructors $\mathsf{F_a}$ do not contain the recursive variable

@@

As explained and exemplified in Section~\ref{sec:GADT-nested},
that assumption is required to make induction work.

@@

\item Case $\mathsf{F\,G\,\ol{b} = H (\ol{F_c\,G\,\ol{b}})}$

@@

That means that $\mathsf{H}$ cannot be a GADT,
as GADTs do not have functorial semantics~\cite{HaskellPaper}
and we would incur again in the same issue encountered in Section~\ref{sec:GADT-nested},

@@
